FraCaS: Temporal Analysis

### 2020-06-09 paper2020/paper.tex
blip
fix a bunch of citations

The semantics of tense and aspect has been a long standing issue in
the study of formal semantics since the early days of Montague Grammar
and a number of different ideas have been put forth on the table to
deal with them throughout the
years~\citep{dowty:2012,prior:2003,steedman:2000,higginbotham:2009,fernando:2015}. The
semantics of tense and aspect has been also considered in the study of
Natural Language Inference (NLI)\todo{citation?}. The various datasets
for NLI that have been proposed by the years contain examples that
have some implicit or explicit reliance on inferences related to tense
and aspect. One of the early datasets used to test logical approaches,
the FraCaS test suite~\citep{cooper:1996} contains a whole section
dedicated to temporal and aspectual inference (section 7). This part
of the FraCaS test suite has been notoriously difficult to tackle. In
fact, all the computational systems ran on the FraCaS test suite so
far skip this section. Additionally, they also skip the anaphora
and ellipsis sections, with the exception of the system presented by
\citet{bernardy_type_2017,bernardy_wide_2019}, which offers a complete whole run of the
FraCaS test suite, excluding the temporal section. In this
paper, we take up the challenge of providing a computationally viable
account of tense and aspect to deal with the notorious section 7\todo{or 7? 8 was not discussed previously} of
the FraCaS test suite. Our account is not meant to be a theoretically
full account of tense and aspect, but rather an account that is driven by the
need to cover the test suite in a way that is general enough to
capture the test suite examples. The account is evaluated on the
entailment properties of various temporal and aspectual examples, as
given by the test suite. As such, we are not getting into the
discussion of how tense and aspect might affect grammaticality or
infelicitousness of various sentences. We take the sentences of the
FraCaS suite to be syntactically correct and the entailment patterns they
are annotated with valid.

@@

\paragraph{GF}
The grammatical framework (GF, \citet{Ranta:GF}) is a powerful parser
generator for natural languages, based on type-theoretical abstract grammars.
This work relies a syntactic representation of the FraCaS test suite in GF
abstract syntax~\citep{Ljunglof:2012}. Thus, we skip the parsing phase
and avoid any syntactic ambiguity.

@@

We support additionally the basic constructions and phenomena present
in the testsuite, including adjectives, adverbs, nouns, verbs,
anaphora, etc. We direct the reader to our previous work for details
\ifanon (citations anonymized) \else
\citet{bernardy_type_2017} \fi, but the particular
treatment of such phenomena is essentially independent from our
treatment of time:
in this paper we simply ignore these aspects beyond the fact that they
are handled correctly in most of the cases of the FraCaS testsuite.

\paragraph{Dynamic semantics}
We additionally assume support for (non-temporal) anaphora, using a
monadic-based dynamic semantics
\cite{unger_dynamic_2011,charlow_monadic_2015,charlow_monadic_2017}.
Our particular implementation is not perfect \ifanon (citations
anonymized) \else \citet{bernardy_wide_2019} \fi but non-temporal
anaphoroi in the testsuite are generally resolved as they should be.

@@

\section{Our Treatment of time}
\label{sec:our-semantics}

@@

This temporal context is an \emph{optional} timespan. That is, it can
be an timespan or an explicitly unspecified context.
\newcommand\nospan{\ensuremath{-}}
\newcommand\anyspan{\ensuremath{\_}}\todo{better notation} The timespan in the context is
optional because, in certain situations, the semantics is different
depending on whether a timespan has been specified externally or not,
as we explain below. A non-present timespan will be represented as
\nospan{}. If a semantic function does not depend on the temporal
context at all, we will write \anyspan{} instead.

@@

\[\constant{unicity}_P : P (t_1,t_2) → P (t_3,t_4) → (t_1 = t_3) ∧ (t_2 = t_4)\]

@@

\providecommand\ncases[1]{{\ensuremath{^{#1}}}}
\begin{table}[hbt]
  \centering
  \small
\begin{tabularx}{\columnwidth}{Xr@{\,\,}r@{\,\,}r@{\,\,}r@{\,\,}r@{\,\,}r@{\,\,}r}
Section      & {\kern -2em} \#FraCaS
                          & This        & FC2         & FC & MINE & Nut  & LP  \\ \hline
Quantifiers  & 75         & .91        & .96         & .96    & .77  & .53  & .93  \\
             &            & \ncases{74}& \ncases{74} &        &      &      &     \ncases{44} \\
Plurals      & 33         & .64        & .82         & .76    & .67  & .52  & .73 \\
             &            &            &             &     &   &   & \ncases{24} \\
Anaphora     & 28         & .79        & .86         &   -    & -    & -    &  -       \\
Ellipsis     & 52         & .81        & .87         &   -    & -    & -    &  -       \\
Adjectives   & 22         & .95        & .95         & .95    & .68  & .32  & .73 \\
             &            & \ncases{20}&  \ncases{20}&     &   &   &  \ncases{12} \\
Comparatives & 31         & .64        & .87         & .56    & .48  & .45  &  -       \\
Temporal     & 75         & .71        &  -          &   -    &   -  &  -   &  -       \\
Verbs        & 8          & .75        & .75         &   -    & -    & -    &  -       \\
Attitudes    & 13         & .85        & .92         & .85    & .77  & .46  & .92  \\ 
             &            &            &             &        &      &      & \ncases {9}  \\ \hline
Total        & 337        & .78        & .89         & .83    & .69  & .50  & .85  \\
             &            & \ncases{329}& \ncases{259}& \ncases{174}  & \ncases{174}& \ncases{174}& \ncases{89}
  \end{tabularx}
  \caption{Accuracy of our system compared to others.
    ``Ours" refers to the approach presented in this paper. When a
    system does not handle the nominal number of test cases (shown in
    the second column), the actual number of test cases attempted is
    shown below the accuracy figure, in smaller font.  ``FraCoq''
    refers to the work of \citet{bernardy:2017}. ``MINE" refers
    to the approach of \citet{Mineshima:2015}, ``NUT" to the CCG
    system that utilizes the first-order automated theorem prover
    \textit{nutcracker} \cite{bos:2008}, and ``Langpro"
    to the system presented by \citet{Abzianidze:2015}. A dash
    indicates that no attempt was made for the section. }
  \label{tab:results}
\end{table}
\paragraph{interaction with negation and group readings}
One reason explaining the lower performance of our systems on some
sections of the testsuite is that our interpretation of time interacts
imperfectly with negation and group readings.

The issue is that the quantification introduced by the temporal marker
is outside the negation. For example, the premiss of (038) is
interpreted as ``there exists a time in the past when every delegate
did not finish the report'' --- which is incorrect in this context.
\fracasex{038}{
\item	No delegate finished the report.
\hyp 	Some delegate finished the report on time.
}
%
We stress that such interaction between quantifiers and negation
already occurs without temporal analysis: one must make a difference
between narrow scoping and wide scoping \cite{todo}. Unfortunately,
the montagovian system over which we base our implementation does not
support this phenomena. (In fact the FraCaS test suite does not appear to
test if the difference between narrow and wide scoping is handled properly.)
We propose that the situation becomes
quantitatively more difficult to handle with the added complication of
time-quantification, but not fundamentally different.
%
A similar issue occurs with group readings, which are interpreted as
conjunctions. Here, the problem is to solve the order of scoping
between this conjunction and time-interval quantifiers.

>>>>

The semantics of tense and aspect has been a long standing issue in
the study of formal semantics since the early days of Montague Grammar
and a number of different ideas have been put forth on the table to
deal with them throughout the
years~\citep{dowty:2012,prior:2003,steedman:2000,higginbotham:2009,fernando:2015}. The
semantics of tense and aspect has been also considered in the study of
Natural Language Inference (NLI). The various datasets
for NLI that have been proposed by the years contain examples that
have some implicit or explicit reliance on inferences related to tense
and aspect. One of the early datasets used to test logical approaches,
the FraCaS test suite~\citep{cooper:1996} contains a whole section
dedicated to temporal and aspectual inference (section 7). This part
of the FraCaS test suite has been notoriously difficult to tackle. In
fact, all the computational systems ran on the FraCaS test suite so
far skip this section. Additionally, they also skip the anaphora
and ellipsis sections, with the exception of the system presented by
\citet{bernardy_type-theoretical_2017,bernardy_wide-coverage_2019}, which offers a complete whole run of the
FraCaS test suite, excluding the temporal section. In this
paper, we take up the challenge of providing a computationally viable
account of tense and aspect to deal with the notorious section 7 of
the FraCaS test suite. Our account is not meant to be a theoretically
full account of tense and aspect, but rather an account that is driven by the
need to cover the test suite in a way that is general enough to
capture the test suite examples. The account is evaluated on the
entailment properties of various temporal and aspectual examples, as
given by the test suite. As such, we are not getting into the
discussion of how tense and aspect might affect grammaticality or
infelicitousness of various sentences. We take the sentences of the
FraCaS suite to be syntactically correct and the entailment patterns they
are annotated with valid.

@@

\paragraph{GF}
The grammatical framework (GF, \citet{ranta_grammatical_2004}) is a powerful parser
generator for natural languages, based on type-theoretical abstract grammars.
This work relies a syntactic representation of the FraCaS test suite in GF
abstract syntax~\citep{Ljunglof:2012}. Thus, we skip the parsing phase
and avoid any syntactic ambiguity.

@@

We support additionally the basic constructions and phenomena present
in the testsuite, including adjectives, adverbs, nouns, verbs,
anaphora, etc. We direct the reader to our previous work for details
\ifanon (citations anonymized) \else
\citet{bernardy_type-theoretical_2017} \fi, but the particular
treatment of such phenomena is essentially independent from our
treatment of time:
in this paper we simply ignore these aspects beyond the fact that they
are handled correctly in most of the cases of the FraCaS testsuite.

\paragraph{Dynamic semantics}
We additionally assume support for (non-temporal) anaphora, using a
monadic-based dynamic semantics
\cite{unger_dynamic_2011,charlow_monadic_2015,charlow_modular_2017}.
Our particular implementation is not perfect \ifanon (citations
anonymized) \else \citet{bernardy_wide-coverage_2019} \fi but non-temporal
anaphoroi in the testsuite are generally resolved as they should be.

@@

\section{Our Treatment of time}
\todo{JP: add a completely worked out example}
\label{sec:our-semantics}

@@

This temporal context is an \emph{optional} timespan. That is, it can
be an timespan or an explicitly unspecified context.
\newcommand\nospan{\ensuremath{-}}
\newcommand\anyspan{\ensuremath{\mathbf{?}}}\todo{better notation} The timespan in the context is
optional because, in certain situations, the semantics is different
depending on whether a timespan has been specified externally or not,
as we explain below. A non-present timespan will be represented as
\nospan{}. If a semantic function does not depend on the temporal
context at all, we will write \anyspan{} instead.

@@

\begin{multline*}
\constant{unicity}_P : P (t_1,t_2) → P (t_3,t_4) → \\ (t_1 = t_3) ∧ (t_2 = t_4)
\end{multline*}

@@

\providecommand\ncases[1]{{\ensuremath{^{#1}}}}
\begin{table}[hbt]
  \centering
  \small
\begin{tabularx}{\columnwidth}{Xr@{\,\,}r@{\,\,}r@{\,\,}r@{\,\,}r@{\,\,}r@{\,\,}r}
Section      & {\kern -2em} \#FraCaS
                          & This        & FC2         & FC & MINE & Nut  & LP  \\ \hline
Quantifiers  & 75         & .91        & .96         & .96    & .77  & .53  & .93  \\
             &            & \ncases{74}& \ncases{74} &        &      &      &     \ncases{44} \\
Plurals      & 33         & .64        & .82         & .76    & .67  & .52  & .73 \\
             &            &            &             &     &   &   & \ncases{24} \\
Anaphora     & 28         & .79        & .86         &   -    & -    & -    &  -       \\
Ellipsis     & 52         & .81        & .87         &   -    & -    & -    &  -       \\
Adjectives   & 22         & .95        & .95         & .95    & .68  & .32  & .73 \\
             &            & \ncases{20}&  \ncases{20}&     &   &   &  \ncases{12} \\
Comparatives & 31         & .64        & .87         & .56    & .48  & .45  &  -       \\
Temporal     & 75         & .71        &  -          &   -    &   -  &  -   &  -       \\
Verbs        & 8          & .75        & .75         &   -    & -    & -    &  -       \\
Attitudes    & 13         & .85        & .92         & .85    & .77  & .46  & .92  \\ 
             &            &            &             &        &      &      & \ncases {9}  \\ \hline
Total        & 337        & .78        & .89         & .83    & .69  & .50  & .85  \\
             &            & \ncases{329}& \ncases{259}& \ncases{174}  & \ncases{174}& \ncases{174}& \ncases{89}
  \end{tabularx}
  \caption{Accuracy of our system compared to others.
    ``Ours" refers to the approach presented in this paper. When a
    system does not handle the nominal number of test cases (shown in
    the second column), the actual number of test cases attempted is
    shown below the accuracy figure, in smaller font.  ``FraCoq''
    refers to the work of \citet{bernardy_type-theoretical_2017}. ``MINE" refers
    to the approach of \citet{Mineshima:2015}, ``NUT" to the CCG
    system that utilizes the first-order automated theorem prover
    \textit{nutcracker} \cite{bos:2008}, and ``Langpro"
    to the system presented by \citet{Abzianidze:2015}. A dash
    indicates that no attempt was made for the section. }
  \label{tab:results}
\end{table}
\paragraph{interaction with negation and group readings}
One reason explaining the lower performance of our systems on some
sections of the testsuite is that our interpretation of time interacts
imperfectly with negation and group readings.

The issue is that the quantification introduced by the temporal marker
is outside the negation. For example, the premiss of (038) is
interpreted as ``there exists a time in the past when every delegate
did not finish the report'' --- which is incorrect in this context.
\fracasex{038}{
\item	No delegate finished the report.
\hyp 	Some delegate finished the report on time.
}
%
We stress that such interaction between quantifiers and negation
already occurs without temporal analysis: one must make a difference
between narrow scoping and wide scoping \cite{todo}\todo{Stergios?}. Unfortunately,
the montagovian system over which we base our implementation does not
support this phenomena. (In fact the FraCaS test suite does not appear to
test if the difference between narrow and wide scoping is handled properly.)
We propose that the situation becomes
quantitatively more difficult to handle with the added complication of
time-quantification, but not fundamentally different.
%
A similar issue occurs with group readings, which are interpreted as
conjunctions. Here, the problem is to solve the order of scoping
between this conjunction and time-interval quantifiers.
